\chapter{Linear Algebra}

\section{Vector Spaces and Linear Operators}
\label{sec:vector-spaces-and-linear-operators}

	\subsection{Vector Spaces}
	
		\indent \indent Let $V$ be a non-empty set with operation '+', called vector addition. Assume that the pair ($V,\, +$) forms an Abelian group i.e. it satisfies the following Axioms:
		
		\begin{enumerate}[label=($A_\arabic*$), leftmargin=3em]
			\setcounter{enumi}{-1}
			\item \label{ax:A0} For all $u, v \in V$, $u + v \in V$, i.e. $V$ is closed under vector addition.
			\item \label{ax:A1} For all $u, v, w \in V$, $u + (v + w) = (u + v) + w$, that is, vector addition is associative.
			\item \label{ax:A2} For all $u, v \in V$, $u + v = v + u$, i.e. vector addition is commutative.
			\item \label{ax:A3} There is a zero element $0_V \in V$ such that $v + 0_V = v$ for all $v \in V$.
			\item \label{ax:A4} Any $v \in V$ has a negative (or additive inverse) $-v \in V$ such that $-v + v = 0_V$.
		\end{enumerate}
		
		\noindent In addition, let $\mathbb{F}$ be a field (e.g. $\mathbb{F} = \mathbb{R}$) and assume that $\mathbb{F}$ acts on $V$ via a mapping $\cdot$ called scalar multiplication and is required to satisfy the following properties.
		
		\begin{enumerate}[label=($M_\arabic*$), leftmargin=3em]
			\item \label{ax:M1} For all $\mu \in \mathbb{F}$ and $u, v \in V$, $\mu \cdot (u + v) = \mu \cdot u + \mu \cdot v$, that is, scalar multiplication is distributive over vector addition.
			\item \label{ax:M2} For all $\mu, \nu \in \mathbb{F}$ and $v \in V$, $(\mu + \nu) \cdot v = \mu \cdot v + \nu \cdot v$, that is, scalar multiplication is distributive over scalar addition.
			\item \label{ax:M3} For all $\mu, \nu \in \mathbb{F}$ and $v \in V$, $\mu \cdot (\nu \cdot v) = (\mu\nu) \cdot v$, that is, scalar multiplication is associative.
			\item \label{ax:M4} For $1 \in \mathbb{F}$, $1 \cdot v = v$ for all $v \in V$.
		\end{enumerate}
	
		\begin{definition}[Vector Space]
			A vector space over a field $\mathbb{F}$ is a set $V$ along with addition and scalar multiplication on $V$, i.e. $\left(V, +, \cdot \right)$ that obeys all axioms above. Elements of $V$ are vectors and elements of $\mathbb{F}$ are scalars.
		\end{definition}
		
		\noindent The element $0_v \in V$ is called the zero-vector, while the element $0 \in \mathbb{F}$ is called the zero-scalar. Further, subtraction in a vector space is defined as $u + (- v) \; \forall \; u, v \in V$. That is, the sum of an element with an inverse element yields a subtraction.
		
		\begin{proposition}
			\label{prop:vspace-properties}
			Let $V$ be a $\mathbb{F}$-vector space
			\begin{enumerate}[label=\roman*.]
				\item For all $\mu \in \mathbb{F}, \, \mu \cdot 0_v = 0_v$ and for all $v \in V, \, 0 \cdot v = 0_v$
				\item If $\mu \cdot v = 0_v$ and $\mathbb{F} \ni \mu \not= 0$, then $V \ni v = 0_v$.
				\item For all $v \in V$, $(-1) \cdot v = -v$
			\end{enumerate}
		\end{proposition}
		
		\begin{proof} This can be proven using the axioms ($A_0$)-($A_4$) and ($M_1$)-($M_4$).
			\begin{enumerate}[label=\roman*.]
				\item From axioms ($A_3$) and ($M_1$), $\mu \cdot 0_V = \mu \cdot (0_V + 0_V) = \mu \cdot 0_V + \mu \cdot 0_V$, and hence, by ($A_3$), $\mu \cdot 0_V = 0_V$. Likewise, from ($M_2$), $0 \cdot v = (0 + 0) \cdot v = 0 \cdot v + 0 \cdot v$, and hence, by ($A_3$), $0 \cdot v = 0_V$.
				\item As $\mathbb{F} \ni \mu \neq 0$, there exists $\mu^{-1} \in \mathbb{F}$ with $\mu^{-1}\mu = 1$. From axioms ($M_4$) and ($M_3$), $v = 1 \cdot v = (\mu^{-1}\mu) \cdot v = \mu^{-1} \cdot (\mu \cdot v)$. Thus, if $\mu \cdot v = 0_V$ we find that $v = \mu^{-1} \cdot 0_V$, and by (i), $v = 0_V$.
				\item From (i) and axioms ($M_2$) and ($M_4$), $0_V = 0 \cdot v = (1 + (-1)) \cdot v = 1 \cdot v + (-1) \cdot v = v + (-1) \cdot v$, and hence $(-1) \cdot v = -v$.
			\end{enumerate}
		\end{proof}
		
		\begin{example}[Matrices]
			The set of all $\left(m \times n\right)$-matrices with entries in $\mathbb{F}$, Mat$(m,n,\mathbb{F})$, is a $\mathbb{F}$-vector space with the usual definitions of adding matrices and multiplying by constants. For $m = n$ we write Mat$(n, \mathbb{F})$.
		\end{example}
		
	
	\subsection{Vector Subspaces}
		
		\indent \indent A non-empty subset $U \subseteq V$ of a vector space need not be a vector space itself, since in general it is not closed under vector addition/scalar multiplication. For example, for any $u_1, u_2 \in U$ it is not necessarily true that $u_1 + u_2 \in U$.
		
		\begin{definition}[Vector Subspace]
			Let $V$ be a $\mathbb{F}$-vector space. A vector subspace of $V$ is a non-empty subset $U$ of $V$ such that $U$ is a $\mathbb{F}$-vector space under the same operations as in $V$.
		\end{definition}
		
		\noindent If $U$ is a subset of a vector space $V$, then axioms \ref{ax:A1}, \ref{ax:A2} and \ref{ax:M1}–\ref{ax:M4} automatically hold in $U$ because they hold throughout $V$.
		
		\begin{definition}[Linear Combination]
			Let $V$ be a $\mathbb{F}$-vector space. A linear combination of the vectors $v_1, \dots, v_m \in V$ is a vector of the form 
			\[
				v = \sum_{i=1}^{m} \mu_i v_i \in V \qquad \text{with} \quad \mu_1, \dots \mu_m \in \mathbb{F}
			\]
		\end{definition}
		
		\begin{proposition}[Subspace Test]
			Let $V$ be a $\mathbb{F}$-vector space and let $U$ be a non-empty subset of $V$. Then $U$ is a vector subspace of $V$ if and only if 
			\[
				\mu_1 u_1 + \mu_2 u_2 \in U \qquad \forall \: u_1, u_2 \in U \; \textnormal{and scalars} \; \mu_1, \mu_2 \in \mathbb{F}
			\]
		\end{proposition}
		
		\noindent Thus, to decide whether $U$ is a vector subspace we do not have to check all the axioms. It is enough to show that $U$ is closed under linear combinations and that it is non-empty. \newline
		\indent Non-emptiness can be shown by demonstrating that the zero-vector is contained in $U$ (note also \ref{ax:A3}).
		

	\subsection{Linear Independence}
		
		\begin{definition}[Span]
			\label{def:span}
			Let $V$ be a $\mathbb{F}$-vector space and $v_1, \dots, v_m \in V$. The span of the set $\{v_1, \dots, v_m\}$ is the set of all possible linear combinations of $v_1, \dots, v_m$, that is,
			\[
				\operatorname{span}_{\mathbb{F}} \{ v_1, \dots, v_m \} := \left\{ \sum_{i=1}^{m} \mu_i v_i \; \middle| \; \mu_1, \dots, \mu_m \in \mathbb{F} \right\}.
			\]
			If $V = \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_m\}$, then $\{v_1, \dots, v_m\}$ is said to be a spanning set for the vector space $V$.
		\end{definition}
		
		\noindent We also note that the zero-vector $0 \in V$ is always contained in the span of any vectors $v_1, \dots, v_m \in V$ since by Proposition~\ref{prop:vspace-properties}, we have $0 = 0v_1 + \dots + 0v_m \in \operatorname{span}_{\mathbb{F}} \{ v_1, \dots, v_m \}$
		
		\begin{example}[Span in $\mathbb{R}^3$]
			In $\mathbb{R}^3$,
			\[
			\operatorname{span}_{\mathbb{R}}\left\{ (1,0,0)^T, (0,1,0)^T \right\} = \left\{ (x,y,0)^T \in \mathbb{R}^3 \;\middle|\; x, y \in \mathbb{R} \right\}
			\]
			describes the $(x,y)$-plane, which is a special instance of a two-dimensional plane in $\mathbb{R}^3$ through the ``origin'' $0 \in \mathbb{R}^3$.
		\end{example}
	
		\noindent Now, let $V$ be a $\mathbb{F}$-vector space and let $v_1, \dots, v_m \in V$.
		
		\begin{definition}[Linear Dependence]
			\label{def:lin-dep}
			The vectors $v_1, \dots, v_m$ are said to be linearly dependent if and only if at least one of the vectors is a linear combination of the others.
		\end{definition} 
		
		\begin{definition}[Linear Independence]
			\label{def:lin-indep}
			The vectors $v_1, \dots, v_m$ are said to be linearly independent if and only if they are not linearly dependent.
		\end{definition}
		
		\begin{remark} 
			If a set $\{v_1, \dots, v_m\}$ contains the zero-vector $0_V \in V$, then the set is linearly dependent since $\mu \cdot 0_V = 0_V$ for all $\mu \in \mathbb{F}$ by Proposition~\ref{prop:vspace-properties}.
		\end{remark}
		
		\begin{proposition}[Extension Lemma]
			\label{prop:extension-lemma}
			Let $\{v_1, \dots, v_m\}$ be a linearly independent set of vectors in a $\mathbb{F}$-vector space $V$. Then, for any $v \in V$,
			$v \in \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_m\} \iff \{v, v_1, \dots, v_m\}$ is a linearly dependent set of vectors.
		\end{proposition}
		
		\begin{proof}
			($\Rightarrow$) Suppose $v \in \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_m\}$. Then $v = \sum_{i=1}^{m} \mu_i v_i $. By Definition~\ref{def:lin-dep}, $v$ is linearly dependent.
			
			($\Leftarrow$) Suppose $\{v, v_1, \dots, v_m\}$ is linearly dependent.
			
			\textit{Case 1:} $v$ is a linear combination of $v_1, \dots, v_m$. Then by the definition of span, $v \in \operatorname{span}_{\mathbb{F}}\{v_1,\dots,v_m\}$.
	
			\textit{Case 2:} some $v_i$ is a linear combination of the other vectors in $\{v, v_1, \dots, v_m\}$, i.e. $v_i = av + \sum_{j \neq i} b_j v_j$. If the coefficient on $v$ were zero, then $v_i$ becomes just a combination of the other $v_j$'s, contradicting linear independence. Hence that coefficient is non-zero, and solving for $v$ gives $v = a^{-1}v_i - \sum_{j\neq i} a^{-1}b_j v_j$, hence $v \in \operatorname{span}_{\mathbb{F}}\{v_1,\dots,v_m\}$.
		\end{proof}
			
		\noindent By contraposition, we now have that $V \ni v \notin \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_m\}$ if and only if $\{v, v_1, \dots, v_m\}$ forms a linearly independent set of vectors.
		
		\begin{definition}[Dimension]
			A $\mathbb{F}$-vector space $V$ is said to be $n$-dimensional if and only if there is a linearly independent set of $n$ vectors and every set consisting of $n+1$ vectors is linearly dependent. $\dim_{\mathbb{F}}(V) = n$ expresses the dimension of $V$. Furthermore, a vector space is said to be of infinite dimension if and only if there is a linearly independent set of vectors of arbitrarily large order.
		\end{definition}
		
	
	\subsection{Bases of Vector Spaces}
		
		\indent \indent We now seek to represent any vector $v \in V$ as a linear combination of a "minimal" set of vectors, also called a basis.
		
		\begin{definition}[Basis]
			Let $V$ be an n-dimensional $\mathbb{F}$-vector space. Any linear independent set of n vectors in $V$ is called a \textit{basis} of $V$.
		\end{definition}
		
		\begin{proposition}
			\label{prop:v-basis}
			Let $V$ be a $\mathbb{F}$-vector space. Any linearly independent spanning set for $V$ is a basis of $V$ and vice versa.
		\end{proposition}
		
		\begin{proof}
			($\Rightarrow$) Let $\{v_1, \dots, v_n\}$ be a spanning set for $V$. Hence, any $v \in V$ lies in the span of $\{v_1, \dots, v_n\}$, that is, $v \in \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_n\}$. If, in addition, $\{v_1, \dots, v_n\}$ is linearly independent, then Proposition~\ref{prop:extension-lemma} implies that $\{v, v_1, \dots, v_n\}$ is a linearly dependent set. Thus, $\dim_{\mathbb{F}}(V) = n$ and so $\{v_1, \dots, v_n\}$ is a basis.
			
			($\Leftarrow$) Conversely, let $n = \dim_{\mathbb{F}}(V)$ and let $\{v_1, \dots, v_n\}$ be a basis of $V$. Then, for any $v \in V$, the set $\{v, v_1, \dots, v_n\}$ is linearly dependent by the definition of a basis. Finally, Proposition~\ref{prop:extension-lemma} implies that $v \in \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_n\}$, which concludes the proof.
		\end{proof}
		
		\begin{example}[Standard Basis for Mat$(m,n,\mathbb{F})$]
			The vector space Mat$(m,n,\mathbb{F})$ is spanned by the matrices $E_{ij}$ having $1$ in the $(i,j)$ position and zero elsewhere. The set of all such matrices is easily shown to be linearly independent. It is the standard basis for Mat$(m,n,\mathbb{F})$ and $\dim_{\mathbb{F}}(\text{Mat}(m,n,\mathbb{F})) = mn$.
		\end{example}
		
		\begin{theorem}[Uniqueness of Representation]
			\label{thm:unique-repr}
			Let $V$ be a $\mathbb{F}$-vector space equipped with a basis. Any vector $v \in V$ can be represented uniquely as a linear combination of the given basis vectors with coefficients in $\mathbb{F}$.
		\end{theorem}
		
		\begin{proof}
			Let $n = \dim_{\mathbb{F}}(V)$ and let $\{v_1, \dots, v_n\}$ be a basis of $V$. By virtue of Proposition~\ref{prop:v-basis}, $V = \operatorname{span}_{\mathbb{F}}\{v_1, \dots, v_n\}$, and thus any $v \in V$ can be written as $v = \sum_{i=1}^{n} \mu_i v_i$ for some $\mu_1, \dots, \mu_n \in \mathbb{F}$. To show uniqueness, let $v = \sum_{i=1}^{n} \mu_i v_i$ and $v = \sum_{i=1}^{n} \nu_i v_i$ be any two representations of $v \in V$ in terms of the basis vectors. Then, equating both expressions, we find that $\sum_{i=1}^{n} (\mu_i - \nu_i) v_i = 0$. Since the basis vectors are linearly independent, all the coefficients must vanish, that is, $\mu_i = \nu_i$ for all $i \in \{1, \dots, n\}$.
		\end{proof}
		
	
	\subsection{Linear Operators}
		
		\begin{definition}[Linear Operator]
			Let $V$ and $W$ be two $\mathbb{F}$-vector spaces. A linear operator is a mapping $T : V \to W$ with $T(\mu u + \nu v) = \mu T(u) + \nu T(v)$ for all $\mu, \nu \in \mathbb{F}$ and $u, v \in V$. Here, $V$ is called the domain of $T$, $W$ the codomain, and $T(v)$ the image of $v \in V$ under $T$.
		\end{definition}
		
		\noindent Linear operators are also called linear transformations, linear maps or homomorphisms. The set of all linear operators from $V$ to $W$ is denoted by $\operatorname{Hom}_{\mathbb{F}}(V,W)$. For any $T_1, T_2 \in \operatorname{Hom}_{\mathbb{F}}(V,W)$, we define 
		\[
			(\mu_1 T_1 + \mu_2 T_2)(v) := \mu_1 T_1(v) + \mu_2 T_2(v) \quad \text{for all} \: v \in V \: \text{and} \: \mu_1, \mu_2 \in \mathbb{F}
		\]
		
		\begin{example}[Transposition as a Linear Operator]
			Let $T : \text{Mat}(m,n,\mathbb{F}) \to \text{Mat}(n,m,\mathbb{F})$ be given by transposition $T : A \mapsto A^T$. For any $A, B \in \text{Mat}(m,n,\mathbb{F})$ and $a, b \in \mathbb{F}$, we have
			\(
				T(aA + bB) = (aA + bB)^T = aA^T + bB^T = aT(A) + bT(B),
			\)
			so $T$ is a linear operator.
		\end{example}
		
		\begin{definition}[Kernel and Image]
			Let $T \in \operatorname{Hom}_{F}(V,W)$. The \textbf{kernel} of $T$ is the set $\ker(T) := \left\{ v \in V | T(v) = 0 \right\} \subseteq V$. The \textbf{image} of T is the set $\operatorname{im}(T) := \left\{ T(v) \in W | v \in V \right\} \subseteq W$.
		\end{definition}
		
		\begin{remark}
			Let $T : V \to W$ be a linear map between $\mathbb{F}$-vector spaces. Then:
			\begin{enumerate}[label=\roman*.]
				\item $\ker(T)=\{v\in V : T(v)=0_W\}$ is a subspace of $V$.
				\item $\operatorname{im}(T)=\{T(v):v\in V\}\subseteq W$ is a subspace of $W$.
			\end{enumerate}
		\end{remark}
		
		\begin{theorem}[Rank-Nullity Theorem]
			\label{thm:rank-nullity}
			For $T \in \operatorname{Hom}_{\mathbb{F}}(V,W)$, we have
			\[
				\dim_{\mathbb{F}}(\operatorname{im}(T)) + \dim_{\mathbb{F}}(\ker(T)) = \dim_{\mathbb{F}}(V).
			\]
		\end{theorem}
		
		\begin{proof}
			Set $n := \dim_{\mathbb{F}}(V)$ and $d := \dim_{\mathbb{F}}(\ker(T))$. Let $\{v_1,\dots,v_d\}$ be a basis of $\ker(T)$. Extend this to a basis $\{v_1,\dots,v_d,u_{d+1},\dots,u_n\}$ of $V$.
			
			We show that $\{T(u_{d+1}),\dots,T(u_n)\}$ is a basis of $\operatorname{im}(T)$. This suffices, since it gives
			\[
				\dim_{\mathbb{F}}(\operatorname{im}(T)) = n-d = \dim_{\mathbb{F}}(V) - \dim_{\mathbb{F}}(\ker(T)).
			\]
			
			\textbf{Linear independence.} Suppose
			\[
				0 = \sum_{i=1}^{n-d}\mu_i\, T(u_{d+i}) = T\left(\sum_{i=1}^{n-d}\mu_i u_{d+i}\right).
			\]
			Then $\sum_{i=1}^{n-d}\mu_i u_{d+i} \in \ker(T)$, so it can be written in terms of the basis $\{v_1,\dots,v_d\}$:
			\[
				\sum_{i=1}^{n-d}\mu_i u_{d+i} = \sum_{j=1}^{d}\nu_j v_j
				\quad\Longleftrightarrow\quad
				\sum_{i=1}^{n-d}\mu_i u_{d+i} - \sum_{j=1}^{d}\nu_j v_j = 0.
			\]
			Since $\{v_1,\dots,v_d,u_{d+1},\dots,u_n\}$ is a basis of $V$, these vectors are linearly independent. Hence all coefficients $\mu_i$ and $\nu_j$ vanish. In particular $\mu_i=0$ for all $i$, so $\{T(u_{d+1}),\dots,T(u_n)\}$ is linearly independent.
			
			\textbf{Spanning.} Let $v\in V$. Write
			\[
				v = \sum_{j=1}^{d}\alpha_j v_j + \sum_{i=1}^{n-d}\beta_i u_{d+i}.
			\]
			Applying $T$ and using $T(v_j)=0$ for all $j$ (since $v_j\in\ker(T)$):
			\[
				T(v) = \sum_{i=1}^{n-d}\beta_i\, T(u_{d+i}).
			\]
			So every element of $\operatorname{im}(T)$ is a linear combination of $\{T(u_{d+1}),\dots,T(u_n)\}$; these vectors span $\operatorname{im}(T)$.
			
			Hence $\{T(u_{d+1}),\dots,T(u_n)\}$ is a basis of $\operatorname{im}(T)$, completing the proof.
		\end{proof}
	
	
\section{The LU Decomposition}
\label{sec:lu-decomposition}
	\subsection{Gaussian Elimination}
		
		\indent \indent Consider a system of $m$ linear equations in $n$ unknowns $x_1, x_2, \dots, x_n$, written in matrix form $Ax = b$, or explicitly, as:
	
		\begin{equation*}
			\begin{bmatrix}
				A_{11} & A_{12} & \dots & A_{1n} \\
				A_{21} & A_{22} & \dots & A_{2n} \\
				\vdots & \vdots & \ddots & \vdots \\
				A_{m1} & A_{m2} & \dots & A_{mn} \\
			\end{bmatrix}
			\begin{bmatrix}
				x_1 \\
				x_2 \\
				\vdots \\
				x_n \\
			\end{bmatrix}
			=
			\begin{bmatrix}
				b_1 \\
				b_2 \\
				\vdots \\
				b_n \\
			\end{bmatrix}
		\end{equation*}
		
		\noindent The key idea of Gaussian elimination is to transform a system of linear equations into a simpler system, while keeping fixed the set of solutions. This leads to the following definition.
		
		\begin{definition}[Elementary Row Operations]
			The following operations are called elementary row operations on a matrix $A$:
			\begin{enumerate}[label=\roman*.]
				\item Interchanging two rows.
				\item Multiplying a row by a non-zero scalar.
				\item Adding a multiple of one row to another row.
			\end{enumerate}
			A matrix $A$ is row-equivalent to another matrix $A'$ if and only if it can be transformed to $A'$ via elementary row operations. 
		\end{definition}
		
		\noindent Gaussian elimination is the repeated application of elementary row operations, mainly used to zero out entries below each pivot.
		
		\begin{definition}[Pivot]
			The leading/first non-zero entry in a row.
		\end{definition}
	
	\subsection{The LU Factorisation}
	
		\noindent It is often convenient to decompose a matrix $A$ into a product of two other matrices. This is called a $LU$ decomposition. We restrict our attention to square, invertible matrices $A \in \text{Mat}(n, \mathbb{F})$, since this is the setting in which the system $Ax = b$ has a unique solution.
	
		\begin{definition}[LU Decomposition]
			Let $A \in \text{Mat}(n, \mathbb{F})$. An LU decomposition of $A$ is a factorisation
			\[
				A = LU
			\]
			where $L \in \text{Mat}(n, \mathbb{F})$ is unit lower triangular (that is, lower triangular with $1$'s on the diagonal) and $U \in \text{Mat}(n, \mathbb{F})$ is upper triangular.
		\end{definition}
	
		\begin{definition}[Leading Principal Minor]
			Let $A \in \text{Mat}(n, \mathbb{F})$. For $k = 1, \dots, n$, the $k$-th leading principal minor of $A$, denoted $\Delta_k$, is the determinant of the $k \times k$ submatrix formed by the first $k$ rows and first $k$ columns of $A$, that is,
			\[
				\Delta_k := \det \begin{bmatrix} 
					A_{11} & \dots & A_{1k} \\ 
					\vdots & \ddots & \vdots \\ 
					A_{k1} & \dots & A_{kk} 
					\end{bmatrix}
			\]
		\end{definition}
	
		\begin{theorem}[Existence and Uniqueness of the LU Decomposition]
			\label{thm:lu-existence}
			Let $A \in \text{Mat}(n, \mathbb{F})$ be invertible. Then $A$ admits a unique LU decomposition if and only if every leading principal minor of $A$ is non-zero, that is, $\Delta_k \neq 0$ for all $k = 1, \dots, n$.
		\end{theorem}
	
		\begin{proof}
			A full treatment of this proof is given in \cite{GolubVanLoan2013}.
		\end{proof}
		
		\begin{example}[A Zero Pivot]
			\label{ex:zero-pivot}
			Consider the matrix
			\[
				A = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}
			\]
			This matrix is invertible, since $\det(A) = -1 \neq 0$. However, its leading principal minors are
			\[
				\Delta_1 = A_{11} = 0, \qquad \Delta_2 = \det(A) = -1,
			\]
			and since $\Delta_1 = 0$, by Theorem~\ref{thm:lu-existence} there does not exist a unique LU decomposition for $A$.
		\end{example}
		
		\begin{definition}[Permutation Matrix]
			\label{def:permutation-matrix}
			A matrix $P \in \mathrm{Mat}(n, \mathbb{F})$ is called a \emph{permutation matrix} if it is obtained from the identity matrix $\mathbb{1}$ by permuting its rows. That is, there exists a permutation $\sigma \in S_n$ such that
			\[
				P_{ij} = \begin{cases} 
							1 & \text{if } j = \sigma(i) \\ 
							0 & \text{otherwise}
						\end{cases}
			\]
			Equivalently, each row and each column of $P$ contains exactly one entry equal to $1$, with all other entries $0$.
		\end{definition}
		
		\noindent To extend the definition of a $LU$ decomposition to any invertible square matrix, we use the permutation matrix $P$ that acts on $A$. This gives the following extension of Theorem~\ref{thm:lu-existence}.
		
		\begin{theorem}[Existence of the $PA=LU$ Decomposition]
			\label{thm:pa-lu-existence}
			Let $A \in \mathrm{Mat}(n, \mathbb{F})$ be invertible. Then there exists a permutation matrix $P \in \mathrm{Mat}(n, \mathbb{F})$ such that $PA$ has all leading principal minors non-zero, and hence, by Theorem~\ref{thm:lu-existence}, a unique decomposition
			\[
				PA = LU,
			\]
			where $L$ is unit lower triangular and $U$ is upper triangular.
		\end{theorem}
		\begin{proof}
			A full treatment of this proof is given in \cite{GolubVanLoan2013}.
		\end{proof}
	
	\subsection{Solving the Linear System}
		We now seek to solve the system of linear equations $Ax=b$ via a $LU$ decomposition.
		
		\begin{proposition}[Solving $Ax=b$ via $PA=LU$]
			\label{prop:solve-pa-lu}
			Let $A \in \mathrm{Mat}(n,\mathbb{F})$ be invertible, let $P$ be a permutation matrix and $L,U$ unit-lower and upper triangular matrices respectively such that $PA = LU$ (Theorem~\ref{thm:pa-lu-existence}), and let $b \in \mathbb{F}^n$. Then the system
			\[
				Ax = b
			\]
			has a unique solution $x$, obtained by first solving $Ly = Pb$ for $y$ by forward substitution, then solving $Ux = y$ for $x$ by back substitution.
		\end{proposition}
		
		\begin{proof}
			$Ax=b$ is equivalent to
			\[
				PAx = Pb.
			\]
			Substituting $PA = LU$,
			\[
				LUx = Pb.
			\]
			Set $y := Ux$. Then the system splits into two triangular systems:
			\[
				Ly = Pb, \qquad Ux = y.
			\]
			Since $L$ is unit lower triangular, $L_{kk}=1\neq 0$ for all $k$, so $Ly=Pb$ has a unique solution obtained by forward substitution:
			\[
				y_k = (Pb)_k - \sum_{j=1}^{k-1} L_{kj} y_j, \qquad k = 1,\dots,n.
			\]
			Since $A$ is invertible and $P$ is invertible, $U = L^{-1}PA$ is invertible, hence $U_{kk} \neq 0$ for all $k$ (an upper triangular matrix is invertible if and only if all diagonal entries are non-zero). So $Ux=y$ has a unique solution obtained by back substitution:
			\[
				x_k = \frac{1}{U_{kk}}\left(y_k - \sum_{j=k+1}^{n} U_{kj}x_j\right), \qquad k = n,n-1,\dots,1.
			\]
			Existence and uniqueness of $y$ and then $x$ give existence and uniqueness of a solution to $Ax=b$.
		\end{proof}
		
		\begin{remark}[The $PA=LU$ Solution Method]
			\label{rem:pa-lu-method}
			Theorem~\ref{thm:pa-lu-existence} and Proposition~\ref{prop:solve-pa-lu} combine into a single procedure for solving $Ax=b$:
			\begin{enumerate}[label=\roman*.]
				\item \textbf{Factor once:} compute $P$, $L$, $U$ such that $PA = LU$ (Theorem~\ref{thm:pa-lu-existence}). This step costs $O(n^3)$ operations.
				\item \textbf{Solve twice:} for any $b \in \mathbb{F}^n$, solve $Ly = Pb$ by forward substitution, then $Ux = y$ by back substitution (Proposition~\ref{prop:solve-pa-lu}). Each substitution costs $O(n^2)$.
			\end{enumerate}
			The point of separating factorisation from substitution is that the $O(n^3)$ cost is paid only once per matrix $A$, while each additional right-hand side $b$ costs only $O(n^2)$. This is the practical motivation for the $PA=LU$ approach over recomputing a full solve (e.g. via Cramer's rule or a fresh elimination) for every new $b$.
		\end{remark}
		
		\begin{example}[Solving $Ax=b$ via $PA=LU$]
			\label{ex:solve-pa-lu-nontrivial}
			Let
			\[
				A = \begin{bmatrix} 
						2 & 3 \\ 4 & 7 
					\end{bmatrix}, 
					\qquad b = \begin{bmatrix} 
						5 \\ 11 
					\end{bmatrix}
			\]
			Here $\Delta_1 = 2 \neq 0$ and $\Delta_2 = \det(A) = 14-12 = 2 \neq 0$, so by Theorem~\ref{thm:lu-existence} $A$ admits an LU decomposition directly, with $P = \mathbb{1}$ (no row swap needed).
			
			\textbf{Factorisation.} Eliminate the $(2,1)$ entry using multiplier $m_{21} = A_{21}/A_{11} = 4/2 = 2$:
			\[
				R_2 \leftarrow R_2 - 2R_1 = 
				\begin{bmatrix} 
					4 & 7 
				\end{bmatrix} 
				- 2\begin{bmatrix}
					2 & 3 
				\end{bmatrix}
				= \begin{bmatrix} 
					0 & 1 
				\end{bmatrix}
			\]
			This gives
			\[
				L = \begin{bmatrix}
						1 & 0 \\ 2 & 1 
					\end{bmatrix}, \qquad 
					U = 
					\begin{bmatrix}
						2 & 3 \\ 0 & 1
					\end{bmatrix}
			\]
			Check: $LU = \begin{bmatrix}
							1 & 0 \\ 2 & 1
						\end{bmatrix}
						\begin{bmatrix}
							2 & 3 \\ 0 & 1
						\end{bmatrix} 
						= \begin{bmatrix}
							2 & 3 \\ 4 & 7
						\end{bmatrix} = A$.
			Correct, and since $P=I$, $PA=LU$ holds trivially.
			
			\textbf{Step 1 (forward substitution):} solve $Ly = Pb = b$.
			\[
				y_1 = b_1 = 5 \qquad y_2 = b_2 - L_{21}y_1 = 11 - 2(5) = 1
			\]
			\begin{center}
			So $y = \begin{bmatrix} 
						5 \\ 1 
					\end{bmatrix}$
			\end{center}
	
			\textbf{Step 2 (back substitution):} solve $Ux = y$.
			\[
				x_2 = \frac{y_2}{U_{22}} = \frac{1}{1} = 1, \qquad 	x_1 = \frac{1}{U_{11}}\left(y_1 - U_{12}x_2\right) = \frac{1}{2}(5 - 3(1)) = 1
			\]
			\begin{center}
				So $x = \begin{bmatrix}
					1 \\ 1
				\end{bmatrix}$
			\end{center}
			
			\textbf{Check:} 
			$Ax = 
				\begin{bmatrix}
					2 & 3 \\ 
					4 & 7 
				\end{bmatrix}
				\begin{bmatrix}
					1 \\
					1
				\end{bmatrix} 
				= 
				\begin{bmatrix}
					5 \\ 
					11
				\end{bmatrix} 
				= b$.
			Correct.
		\end{example}
		
		
	\subsection{Partial Pivoting}
		
		\indent \indent The example above shows that the leading-principal-minor condition in Theorem~\ref{thm:lu-existence} can fail in two ways: exactly (Example~\ref{ex:zero-pivot}), or nearly, via a small pivot that leaves the algorithm formally intact but numerically unreliable. \emph{Partial pivoting} is the standard modification to Gaussian elimination that addresses both.
		
		\begin{definition}[Partial Pivoting]
			\label{def:partial-pivoting}
			Let $A \in \mathrm{Mat}(n,\mathbb{F})$. Set $A^{(1)} := A$, and for $k = 2, \dots, n$, let $A^{(k)}$ denote the matrix obtained after the eliminations and row swaps of steps $1,\dots,k-1$ have been applied. At step $k$ of Gaussian elimination, define the \emph{active column} to be the entries $A^{(k)}_{kk}, A^{(k)}_{k+1,k}, \dots, A^{(k)}_{nk}$, i.e.\ column $k$ restricted to rows $k,\dots,n$. Let
			\[
				r := \operatorname*{arg\,max}_{k \le i \le n} 	|A^{(k)}_{ik}|
			\]
			be a row index attaining the largest absolute value in the active column. Partial pivoting swaps row $r$ with row $k$ before proceeding with elimination as normal. Each such swap is a transposition; the composition of these transpositions, in the order performed across steps $k=1,\dots,n-1$, defines the permutation $\sigma \in S_n$ of Definition~\ref{def:permutation-matrix}, and hence the permutation matrix $P$.
		\end{definition}
		
		\noindent With this modification, the substantive question is whether the procedure can always be carried out. That is, whether a valid pivot always exists at every step, for any invertible $A$.
		
		\begin{proposition}[Partial Pivoting on an Invertible Matrix]
			\label{prop:partial-pivoting-exists}
			Let $A \in \mathrm{Mat}(n,\mathbb{F})$ be invertible. Then at every step $k=1,\dots,n-1$ of Gaussian elimination with partial pivoting, the active column contains at least one non-zero entry, so a valid pivot row $r$ always exists.
		\end{proposition}
		\begin{proof}
			A full treatment of this proof is given in \cite{GolubVanLoan2013}.
		\end{proof}
		
		\noindent But why should one bother with partial pivoting? After all, Theorem~\ref{thm:pa-lu-existence} already guarantees some valid $P$ exists for any invertible $A$, so why does the specific pivoting strategy matter?
		
		\begin{proposition}[Bounded Multipliers Under Partial Pivoting]
			\label{prop:bounded-multipliers}
			Suppose partial pivoting is used at every step of Gaussian elimination. Then every multiplier satisfies
			\[
				|m_{ik}| \le 1 \qquad \text{for all } i > k
			\]
		\end{proposition}
		\begin{proof}
			Fix step $k$. By construction, the pivot $A^{(k)}_{kk}$ (after the swap) satisfies
			\[
				|A^{(k)}_{kk}| \ge |A^{(k)}_{ik}| \qquad \text{for 	all } i = k, \dots, n
			\]
			since $A^{(k)}_{kk}$ was chosen to be the entry of largest absolute value among the active column. In particular, for every $i > k$,
			\[
				|m_{ik}| = 	\left|\frac{A^{(k)}_{ik}}{A^{(k)}_{kk}}\right| \le 1
			\]
		\end{proof}
	
	
\section{Inner Product Spaces and QR Decomposition}
\label{sec:inner-products}
	
	\subsection{Inner Product Spaces}
		
		\indent \indent In Section~\ref{sec:vector-spaces-and-linear-operators}, we defined a vector space $V$, but a vector space alone carries no notion of length or angle between vectors. Equipping $V$ with such a notion is the first step toward orthogonality, which in turn is the basis of the QR decomposition developed later in this section.
		
		Recall the dot product for $\mathbb{R}^n$:
		\[
			x \cdot y = \sum_{i=1}^{n} x_i y_i.
		\]
		
		This immediately gives a notion of length via $\|x\| := \sqrt{x \cdot x}$.
		
		The dot product on $\mathbb{R}^n$ has three properties worth isolating: it is symmetric in $x$ and $y$, linear in each argument and positive-definite ($x \cdot x \ge 0$, with equality iff $x=0$). Abstracting exactly these properties to an arbitrary vector space yields the definition of an inner product.
		
		Throughout this section, $V$ will denote a real vector space, i.e.\ $\mathbb{F} = \mathbb{R}$.
		
		
		\begin{definition}[Inner Product / Inner Product Space]
			\label{def:inner-product}
			An inner product on a real vector space $V$ is a map $\langle \cdot, \cdot \rangle : V \times V \to \mathbb{R}$ that satisfies the following properties for all $u, v, w \in V$ and all $\lambda \in \mathbb{R}$:
			
			\begin{enumerate}[label=(IP\arabic*)]
				\item \label{ax:IP1} \textbf{Symmetry:} $\langle u, v \rangle = \langle v, u \rangle$
				\item \label{ax:IP2} \textbf{Linearity in the first argument:} $\langle u + v, w \rangle = \langle u, w \rangle + \langle v, w \rangle$ and $\langle \lambda u, v \rangle = \lambda \langle u, v \rangle$
				\item \label{ax:IP3} \textbf{Positive-definiteness:} $\langle v, v \rangle \ge 0$ for all $v \in V$, with equality $\langle v, v \rangle = 0$ if and only if $v = 0_V$
			\end{enumerate}
			
			The pair $(V, \langle \cdot, \cdot \rangle)$ is called an \textbf{inner product space}.
			
		\end{definition}
		
		\begin{example}[Trace Inner Product]
			Consider $\mathrm{Mat}(n, \mathbb{R})$. Equipped with the trace inner product $\langle x | y \rangle := \operatorname{tr}(A^\top B)$ for all $x, y \in \mathrm{Mat}(n, \mathbb{R})$ the pair $(\mathrm{Mat}(n, \mathbb{R}), \langle \cdot, \cdot \rangle)$ is an inner product space.
		\end{example}
		
		\begin{remark}
			In a real vector space, the properties of symmetry \ref{ax:IP1} and linearity in the first argument \ref{ax:IP2} together imply linearity in the second argument, meaning the inner product is fully bilinear.
		\end{remark} 
		
		
		\begin{definition}[Induced Norm] 
			\label{def:induced-norm}
			Let $(V, \langle \cdot, \cdot \rangle)$ be an inner product space. For each vector $v \in V$, the \textbf{norm} of $v$ induced by the inner product, denoted by $\|v\|$, is defined as:
			\[
				\|v\| := \sqrt{\langle v, v \rangle} 
			\]
		\end{definition}
		
		\begin{remark}
			\label{rem:bilinearity}
			Note that the quantity ${\langle v, v \rangle}$ is non-negative from positive-definiteness \ref{ax:IP3}, hence the induced norm is always well-defined in $\mathbb{R}$.
		\end{remark}
		
		\begin{theorem}[Cauchy-Schwarz Inequality] \label{thm:cauchy-schwarz}
			Let $V$ be an inner product space. For all vectors $u, v \in V$, the following inequality holds:
			\[ 
				|\langle u, v \rangle| \le \|u\| \|v\|
			\]
			This inequality is an equality if and only if $u, v$ are linearly dependent (Definition~\ref{def:lin-dep}).
		\end{theorem}
		
		\begin{proof}
			Let $V$ be a real inner product space. We wish to show that for all $u, v \in V$, $|\langle u, v \rangle| \le \|u\| \|v\|$.
			
			Suppose $v = 0_V$. $0_V = 0 \cdot u$ (Proposition~\ref{prop:vspace-properties}); combined with symmetry \ref{ax:IP1} and linearity in the first argument \ref{ax:IP2}, $\langle u, 0_V \rangle = \langle u, 0 \cdot u \rangle = \langle 0\cdot u, u\rangle = 0\langle u,u\rangle = 0$. The inequality then becomes:
			\[ 
				|\langle u, 0_V \rangle| = 0 \le \|u\| \cdot 0 = 0
			\]
			which holds as an equality. Thus, the theorem is true for $v = 0_V$.
			
			Now assume $v \neq 0_V$. For any real number $t \in \mathbb{R}$, consider the function:
			\[ 
				f(t) = \|u - tv\|^2
			\]
			By the definition of the induced norm, $f(t) = \langle u - tv, u - tv \rangle$. By positive-definiteness \ref{ax:IP3}, the inner product of any vector with itself is non-negative, and is zero iff the vector is the zero vector. Therefore, $f(t) \ge 0$ for all $t \in \mathbb{R}$.
			
			Using bilinearity (Remark~\ref{rem:bilinearity}), we expand the inner product:
			
			\begin{align*}
				f(t) &= \langle u, u - tv \rangle - t\langle v, u - tv \rangle \\
				&= \langle u, u \rangle - t\langle u, v \rangle - t\langle v, u \rangle + t^2\langle v, v \rangle \\
				&= \|u\|^2 - 2t\langle u, v \rangle + t^2\|v\|^2
			\end{align*}
			
			This is a quadratic in $t$ of the form $at^2 + bt + c$, where $a = \|v\|^2$, $b = -2\langle u, v \rangle$, and $c = \|u\|^2$.
			
			Since $v \neq 0_V$, positive-definiteness \ref{ax:IP3} gives $\|v\|^2 = \langle v,v\rangle > 0$, so $\|v\|^2 \neq 0$ and division by it below is valid.
			
			Following the minimisation route, we choose the specific value of $t$ that minimises $f(t)$. For a quadratic $at^2 + bt + c$, the minimum occurs at $t = -b/(2a)$. Here, we set:
			\[
				t_0 = \frac{2\langle u, v \rangle}{2\|v\|^2} = \frac{\langle u, v \rangle}{\|v\|^2}
			\]
			Plugging $t_0$ into our inequality $f(t) \ge 0$:
			
			\begin{align*}
				0 &\le \|u\|^2 - 2\left(\frac{\langle u, v \rangle}{\|v\|^2}\right)\langle u, v \rangle + \left(\frac{\langle u, v \rangle}{\|v\|^2}\right)^2 \|v\|^2 \\
				0 &\le \|u\|^2 - \frac{2\langle u, v \rangle^2}{\|v\|^2} + \frac{\langle u, v \rangle^2}{\|v\|^2} \\
				0 &\le \|u\|^2 - \frac{\langle u, v \rangle^2}{\|v\|^2}
			\end{align*}
			
			Rearranging terms, we find $\frac{\langle u, v \rangle^2}{\|v\|^2} \le \|u\|^2$, which implies $\langle u, v \rangle^2 \le \|u\|^2 \|v\|^2$. Taking the square root of both sides yields the desired inequality:
			\[ 
				|\langle u, v \rangle| \le \|u\| \|v\|
			\]
			
			Equality $|\langle u, v \rangle| = \|u\| \|v\|$ holds iff $f(t_0) = 0$. By positive-definiteness \ref{ax:IP3}, $f(t_0) = \|u - t_0v\|^2 = 0$ iff $u - t_0v = 0_V$. This implies $u = t_0v$, meaning $u$ is a scalar multiple of $v$. Thus, equality holds iff $u$ and $v$ are linearly dependent.
		\end{proof}
		
		\begin{remark}[Angle Between Vectors]
			\label{rem:angle-between-vectors}
			Let $V$ be a real inner product space and let $u, v \in V$ with $u, v \neq 0_V$. By Cauchy-Schwarz (Theorem~\ref{thm:cauchy-schwarz}),
			\[
				|\langle u, v \rangle| \le \|u\|\|v\|
			\]
			Since $u, v \neq 0_V$, positive-definiteness \ref{ax:IP3} gives $\|u\|^2 > 0$ and $\|v\|^2 > 0$, so $\|u\| \neq 0$ and $\|v\| \neq 0$, and we may divide through by $\|u\|\|v\|$:
			\[
				\left| \frac{\langle u,v \rangle}{\|u\|\|v\|} \right| \le 1
			\]
			That is, the ratio $\frac{\langle u, v \rangle}{\|u\|\|v\|}$ always lies in $[-1,1]$, which is exactly the range of $\cos$. Since $\cos : [0,\pi] \to [-1,1]$ is a bijection, there is a unique $\theta \in [0,\pi]$ satisfying
			\[
				\cos\theta = \frac{\langle u,v \rangle}{\|u\|\|v\|}
			\]
			This uniqueness is precisely what licenses calling $\theta$ \textbf{the angle between} $u$ and $v$: without Cauchy-Schwarz bounding the ratio into $[-1,1]$ first, no such $\theta$ would be well-defined at all.
		\end{remark}
		
		\begin{proposition}[Triangle Inequality]
			\label{prop:triangle-inequality}
			Let $V$ be a real inner product space. Then for all $u, v \in V$
			\[
				\|u+v\| \le \|u\| + \|v\|
			\]
			(a consequence of Cauchy-Schwarz, Theorem~\ref{thm:cauchy-schwarz}).
		\end{proposition}
		
		\begin{proof}
			We bound $\|u+v\|^2$ and take square roots at the end.
			
			By the definition of the induced norm and bilinearity (Remark~\ref{rem:bilinearity}),
			\begin{align*}
				\|u+v\|^2 &= \langle u+v, u+v \rangle \\
				&= \langle u,u\rangle + \langle u,v\rangle + \langle v,u\rangle + \langle v,v\rangle \\
				&= \|u\|^2 + 2\langle u,v\rangle + \|v\|^2
			\end{align*}
			using symmetry \ref{ax:IP1} to combine $\langle u,v\rangle$ and $\langle v,u\rangle$.
			
			By Cauchy-Schwarz (Theorem~\ref{thm:cauchy-schwarz}), $|\langle u,v\rangle| \le \|u\|\|v\|$, and in particular $\langle u,v\rangle \le |\langle u,v\rangle| \le \|u\|\|v\|$. Substituting this bound on the cross term,
			\[
				\|u+v\|^2 \le \|u\|^2 + 2\|u\|\|v\| + \|v\|^2
			\]
			The right-hand side is a perfect square:
			\[
				\|u\|^2 + 2\|u\|\|v\| + \|v\|^2 = \left(\|u\| + \|v\|\right)^2
			\]
			So $\|u+v\|^2 \le \left(\|u\|+\|v\|\right)^2$. Since both $\|u+v\|$ and $\|u\|+\|v\|$ are non-negative, taking square roots preserves the inequality:
			\[
				\|u+v\| \le \|u\| + \|v\|
			\]
		\end{proof}
		
	\subsection{Orthogonality}
		
		\noindent The angle between vectors (Remark~\ref{rem:angle-between-vectors}) showed that $\theta = \pi/2$ if and only if $\cos\theta = 0$, which by that same remark holds if and only if $\langle u, v \rangle = 0$. This special case (i.e. an inner product of zero) is important enough to warrant its own name.
		
		\begin{definition}[Orthogonal Set / Orthonormal Set]
			\label{def:orthogonal-orthonormal-set}
			Let $V$ be an inner product space and let $v_1, \dots, v_m \in V$ be a set of vectors.
			\begin{enumerate}[label=\roman*.]
				\item The set $\{v_1,\dots,v_m\}$ is \emph{orthogonal} if $\langle v_i, v_j \rangle = 0$ for all $i \neq j$.
				\item The set $\{v_1,\dots,v_m\}$ is \emph{orthonormal} if it is orthogonal and, in addition, $\|v_i\| = 1$ for all $i$.
			\end{enumerate}
		\end{definition}
		
		\begin{remark}
			Both conditions in Definition~\ref{def:orthogonal-orthonormal-set} compress into a single statement using the Kronecker delta: $\{v_1,\dots,v_m\}$ is orthonormal if and only if
			\[
				\langle v_i, v_j \rangle = \delta_{ij} \qquad \text{for all } i,j
			\]
			where $\delta_{ij} = 1$ if $i=j$ and $0$ otherwise.
		\end{remark}
		
		\begin{proposition}[Orthogonality and Linear Independence]
			\label{prop:orthogonal-and-linearly-independent}
			Let $V$ be an inner product space and let $v_1, \dots, v_m \in V$ be an orthogonal set of nonzero vectors. Then $v_1, \dots, v_m$ are linearly independent.
		\end{proposition}
		
		\begin{proof}
			Suppose
			\[
				c_1 v_1 + \cdots + c_m v_m = 0_V
			\]
			for some scalars $c_1,\dots,c_m \in \mathbb{R}$. We show $c_i = 0$ for all $i$.
			
			Fix $k \in \{1,\dots,m\}$ and take the inner product of both sides with $v_k$:
			\[
				\left\langle \sum_{i=1}^{m} c_i v_i,\; v_k \right\rangle = \langle 0_V, v_k \rangle
			\]
			By symmetry \ref{ax:IP1} and the same argument as in the Cauchy–Schwarz proof, $\langle 0_V, v_k \rangle = \langle v_k, 0_V \rangle = 0$.
			
			Expanding the left-hand side by bilinearity (Remark~\ref{rem:bilinearity}),
			\[
				\sum_{i=1}^{m} c_i \langle v_i, v_k \rangle = 0
			\]
			Since $\{v_1,\dots,v_m\}$ is orthogonal, $\langle v_i, v_k\rangle = 0$ for all $i \neq k$, so every term with $i\neq k$ vanishes, leaving only the $i=k$ term:
			\[
				c_k \langle v_k, v_k \rangle = 0, \qquad \text{i.e.} \qquad c_k \|v_k\|^2 = 0
			\]
			Since $v_k \neq 0_V$, positive-definiteness \ref{ax:IP3} gives $\|v_k\|^2 > 0$, so $\|v_k\|^2 \neq 0$, and hence $c_k = 0$.
			
			Since $k$ was arbitrary, $c_k = 0$ for every $k=1,\dots,m$, so $v_1,\dots,v_m$ are linearly independent.
		\end{proof}
		
		\begin{theorem}[Gram-Schmidt Process]
			\label{thm:gram-schmidt}
			Let $V$ be an inner product space and let $v_1, \dots, v_m \in V$ be linearly independent. Define
			\[
				u_1 = v_1, \qquad u_k = v_k - \sum_{j=1}^{k-1} \frac{\langle v_k, u_j \rangle}{\langle u_j, u_j \rangle} u_j \quad \text{for } k = 2, \dots, m
			\]
			Then $u_1, \dots, u_m$ is an orthogonal, hence linearly independent (Proposition~\ref{prop:orthogonal-and-linearly-independent}), set of nonzero vectors satisfying
			\[
				\operatorname{span}\{u_1, \dots, u_k\} = \operatorname{span}\{v_1, \dots, v_k\} \qquad \text{for every } k = 1, \dots, m
			\]
		\end{theorem}
		
		\begin{proof}
			A full treatment of this proof is given in \cite{GolubVanLoan2013}.
		\end{proof}
		
		\begin{corollary}[Existence of an Orthonormal Basis]
			\label{cor:orthonormal-basis-exists}
			Let $V$ be a finite-dimensional inner product space with $\dim V = n$. Then, by Theorem~\ref{thm:gram-schmidt}, $V$ has an orthonormal basis.
		\end{corollary}
		
		\begin{proof}
			Since $\dim V = n$, $V$ has a basis $v_1,\dots,v_n$.
			
			Apply Gram-Schmidt (Theorem~\ref{thm:gram-schmidt}) to $v_1,\dots,v_n$, producing an orthogonal set $u_1,\dots,u_n$ with
			\[
				\operatorname{span}_{\mathbb{R}}\{u_1,\dots,u_n\} = \operatorname{span}_{\mathbb{R}}\{v_1,\dots,v_n\} = V
			\]
			By Theorem~\ref{thm:gram-schmidt}, $u_1,\dots,u_n$ is an orthogonal, linearly independent set of nonzero vectors, with nonzero vectors. Being $n$ linearly independent vectors spanning the $n$-dimensional space $V$, $u_1,\dots,u_n$ form a basis of $V$ by Proposition~\ref{prop:v-basis}.
			
			Define $e_k := u_k / \|u_k\|$ for each $k=1,\dots,n$. Since $u_k \neq 0_V$, positive-definiteness \ref{ax:IP3} gives $\|u_k\| \neq 0$, so this division is well-defined.
			
			For $i \neq k$, by bilinearity (Remark~\ref{rem:bilinearity}),
			\[
				\langle e_i, e_k \rangle = \left\langle \frac{u_i}{\|u_i\|}, \frac{u_k}{\|u_k\|} \right\rangle = \frac{1}{\|u_i\|\|u_k\|}\langle u_i, u_k \rangle = \frac{1}{\|u_i\|\|u_k\|} \cdot 0 = 0
			\]
			since $\langle u_i, u_k\rangle = 0$ by orthogonality of $u_1,\dots,u_n$, and dividing zero by the nonzero scalar $\|u_i\|\|u_k\|$ leaves zero.
			
			By bilinearity (Remark~\ref{rem:bilinearity}),
			\[
				\|\lambda v\|^2 = \langle \lambda v, \lambda v \rangle = \lambda^2 \langle v,v \rangle = \lambda^2 \|v\|^2
			\]
			so $\|\lambda v\| = |\lambda| \, \|v\|$ (taking non-negative square roots). Applying this with $\lambda = 1/\|u_k\|$,
			\[
				\|e_k\| = \left\| \frac{u_k}{\|u_k\|} \right\| = \frac{1}{\|u_k\|}\|u_k\| = 1
			\]
			
			Hence $\{e_1,\dots,e_n\}$ is orthonormal (Definition~\ref{def:orthogonal-orthonormal-set}). Since scaling each $u_k$ by the nonzero scalar $1/\|u_k\|$ changes neither the span nor linear independence of $\{u_1,\dots,u_n\}$, $\{e_1,\dots,e_n\}$ remains a basis of $V$. Thus $\{e_1,\dots,e_n\}$ is an orthonormal basis of $V$.
		\end{proof}
		
	\subsection{QR Decomposition}
		
		\begin{definition}[QR Decomposition]
			\label{def:qr-decomp}
			Let $A \in \mathrm{Mat}(n, \mathbb{R})$. A QR decomposition of $A$ is a factorisation $A=QR$ where $Q \in \mathrm{Mat}(n,\mathbb{R})$ has orthonormal columns, that is, the columns $q_1, \dots, q_n$ form an orthonormal set (Definition~\ref{def:orthogonal-orthonormal-set}); and $R \in \mathrm{Mat}(n,\mathbb{R})$ is upper triangular with positive diagonal entries.
		\end{definition}
		
		\begin{theorem}[Existence and Uniqueness of the QR Decomposition]
			\label{thm:existence-qr-decomp}
			Let $A \in \mathrm{Mat}(n,\mathbb{R})$ be invertible. Then $A$ admits a QR decomposition $A=QR$ (as in Definition~\ref{def:qr-decomp}), and this decomposition is unique.
		\end{theorem}
		
		\begin{proof}
			Let $a_1,\dots, a_n$ denote the columns of $A$. Since $A$ is invertible, its columns are linearly independent.
			
			Apply Gram-Schmidt (Theorem~\ref{thm:gram-schmidt}) to $a_1,\dots,a_n$ to obtain an orthogonal set $u_1,\dots,u_n$ with
			\[
				\operatorname{span}_{\mathbb{R}}\{u_1,\dots,u_k\} = \operatorname{span}_{\mathbb{R}}\{a_1,\dots,a_k\} \qquad \text{for every } k=1, \dots, n
			\]
			and normalise $q_k := u_k/\|u_k\|$ for each $k$, exactly as in the construction of Corollary~\ref{cor:orthonormal-basis-exists}; we do not re-derive well-definedness or orthonormality of $q_1,\dots,q_n$ here, having already established both.
			
			Let $Q$ be the matrix with columns $q_1,\dots,q_n$. By construction, $Q$ has orthonormal columns.
			
			We now construct $R$. Since $\operatorname{span}_{\mathbb{R}}\{q_1,\dots,q_k\} = \operatorname{span}_{\mathbb{R}}\{u_1,\dots,u_k\} = \operatorname{span}_{\mathbb{R}}\{a_1,\dots,a_k\}$ for every $k$, each $a_k$ lies in $\operatorname{span}_{\mathbb{R}}\{q_1,\dots,q_k\}$, so
			\[
			a_k = \sum_{j=1}^{k} c_j\, q_j
			\]
			for some scalars $c_1,\dots,c_k$. Fix $i \le k$ and take the inner product of both sides with $q_i$. By bilinearity (Remark~\ref{rem:bilinearity}),
			\[
				\langle a_k, q_i \rangle = \sum_{j=1}^{k} c_j \langle q_j, q_i \rangle
			\]
			Since $\{q_1,\dots,q_n\}$ is orthonormal, $\langle q_j,q_i\rangle = 0$ for $j\neq i$ and $\langle q_i,q_i\rangle = 1$, so every term vanishes except $j=i$, leaving $\langle a_k, q_i\rangle = c_i$. As $i \le k$ was arbitrary, this identifies every coefficient: $c_j = \langle a_k, q_j\rangle$ for all $j=1,\dots,k$, so
			\[
				a_k = \sum_{j=1}^{k} \langle a_k, q_j \rangle\, q_j
			\]
			Define $R_{jk} := \langle a_k, q_j \rangle$ for $j \le k$ and $R_{jk} := 0$ for $j > k$. This is upper triangular by construction, and the identity above shows precisely that $a_k = \sum_{j=1}^n R_{jk} q_j$ is the $k$-th column of $QR$, so $A = QR$.
			
			It remains to show $R_{kk} > 0$ for every $k$. Fix $k$. By the Gram-Schmidt relation,
			\[
				u_k = a_k - \sum_{j<k} \operatorname{proj}_{u_j}(a_k)
			\]
			so
			\[
				a_k = u_k + \sum_{j<k} \operatorname{proj}_{u_j}(a_k)
			\]
			Each $\operatorname{proj}_{u_j}(a_k)$ is by definition a scalar multiple of $u_j$, hence of $q_j$. So $a_k = u_k + \sum_{j<k} d_j q_j$ for some scalars $d_j$. Taking the inner product with $q_k$ and using bilinearity (Remark~\ref{rem:bilinearity}),
			\[
				\langle a_k, q_k \rangle = \langle u_k, q_k \rangle + \sum_{j<k} d_j \langle q_j, q_k \rangle
			\]
			Since $\{q_1,\dots,q_n\}$ is orthonormal, $\langle q_j,q_k\rangle = 0$ for $j \neq k$, so every term in the sum vanishes, leaving
			\[
				R_{kk} = \langle a_k, q_k \rangle = \langle u_k, q_k \rangle = \left\langle u_k, \frac{u_k}{\|u_k\|} \right\rangle = \frac{\|u_k\|^2}{\|u_k\|} = \|u_k\|
			\]
			Since $u_k \neq 0_V$ (Gram-Schmidt produces nonzero vectors, as used already in Corollary~\ref{cor:orthonormal-basis-exists}), positive-definiteness \ref{ax:IP3} gives $\|u_k\| > 0$. Hence $R_{kk} > 0$ for all $k$. This establishes existence of the decomposition $A = QR$ with the stated properties.
			
			To prove uniqueness of the QR decomposition, we rely on \cite{GolubVanLoan2013}.
		\end{proof}
		
		\begin{proposition}[Solving $Ax=b$ via $A=QR$]
			\label{prop:solve-qr}
			Let $A \in \mathrm{Mat}(n,\mathbb{R})$ be invertible, let $Q, R$ be such that $A = QR$ (Theorem~\ref{thm:existence-qr-decomp}), and let $b \in \mathbb{R}^n$. Then the system $Ax=b$ has a unique solution $x$, obtained by first computing $c := Q^\top b$, then solving $Rx = c$ for $x$ by back substitution.
		\end{proposition}
		
		\begin{proof}
			Substituting $A = QR$, the system $Ax=b$ becomes
			\[
				QRx = b
			\]
			Since $Q$ has orthonormal columns, $Q^\top Q = I$. As $Q$ is square, this left inverse is automatically a two-sided inverse, so $Q$ is invertible with $Q^{-1} = Q^\top$.
			
			Multiplying both sides of $QRx=b$ on the left by $Q^\top$,
			\[
				Q^\top Q R x = Q^\top b \quad \Longrightarrow \quad Rx = Q^\top b
			\]
			Define $c := Q^\top b$. Unlike the analogous step in the $PA=LU$ method, computing $c$ requires no triangular solve — it is a single direct matrix-vector product.
			
			It remains to solve $Rx = c$. Since $R$ is upper triangular with $R_{kk} > 0$ for all $k$ (Theorem~\ref{thm:existence-qr-decomp}), $R$ is invertible, so $Rx=c$ has a unique solution obtained by back substitution:
			\[
				x_k = \frac{1}{R_{kk}}\left(c_k - \sum_{j=k+1}^{n} R_{kj}x_j\right) \qquad k = n, n-1, \dots, 1
			\]
			Existence and uniqueness of $c$ and then $x$ give existence and uniqueness of a solution to $Ax=b$.
		\end{proof}
		
		

	
	